---
title: "Problem1_coding"
author: "Zihan Dong"
date: '2024-07-03'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
4.
```{r}
#----------------------------------
library(MASS)
set.seed(1)

### training data
n <- 600
p <- 600

# generate Sigma
rho <- 0.3
Sigma <- matrix(rep(0.3, p^2), nrow = p, ncol = p)
diag(Sigma) <- 1
# generate matrix X and vector xi0 (xi_star in the above description)
X <- mvrnorm(n, rep(0,p), Sigma = Sigma)
xi0 <- Sigma %*% c(-1,1,-1,1,1,rep(0, p-5))

# Calculate Sigma hat (X^T X / n)
Sigma_hat <- t(X) %*% X / n
# Define the coordinate descent function
coordinate_descent <- function(X, xi0, lambda, delta = 1e-5, max_iter = 1000) {
  p <- ncol(X)
  u <- rep(0, p)  # Initial value is 0
  u_old <- u
  iter_count <- 0
  u_history <- matrix(0, nrow = max_iter, ncol = p)
  
  soft_threshold <- function(z, gamma) {
    sign(z) * max(abs(z) - gamma, 0)
  }
  
  for (iter in 1:max_iter) {
    u_old <- u
    for (j in 1:p) {
      x_j <- X[, j]
      X_rest <- X[, -j]
      u_rest <- u[-j]
      S_j <- sum(x_j^2)/n
      
      rho_j <- (xi0[j] - x_j %*% X_rest %*% u_rest/n) / S_j
      u[j] <- soft_threshold(rho_j, lambda / S_j)
    }
    
    # Save the current iteration result
    u_history[iter, ] <- u
    
    # Check convergence criteria
    if (sum(abs(u - u_old), na.rm = TRUE) < delta) {
      iter_count <- iter
      u_history <- u_history[1:iter, ]  # Truncate to the actual number of iterations
      break
    }
  }
  
  if (iter == max_iter) {
    iter_count <- max_iter
  }
  
  list(u = u, iter_count = iter_count, u_history = u_history)
}

# Calculate the value of lambda
lambda <- 1.2 * sqrt(log(p) / n)
# Solve using coordinate descent
result <- coordinate_descent(X, xi0, lambda)

# Print results
cat("Number of iterations:", result$iter_count, "\n")
cat("Final iterate (first 10 components):", result$u[1:10], "\n")

# Function to plot the iterations of specific components
plot_iterations <- function(u_history, components) {
  iter_count <- nrow(u_history)
  par(mfrow = c(length(components), 1), mar = c(4, 4, 2, 1))
  
  for (j in components) {
    plot(1:iter_count, u_history[, j], type = "l", main = paste("Component", j),
         xlab = "Iteration", ylab = paste("u[", j, "]", sep = ""))
  }
}

# Plot iterations for components j = 2, 4, 6, 8, 10
plot_iterations(result$u_history, c(2, 4, 6, 8, 10))
```

```{r}
# Check the stationary condition with tolerance
check_stationary_condition <- function(u, Sigma, xi0, lambda, tol = 1e-5) {
  condition_met <- TRUE
  for (j in 1:length(u)) {
    value <-Sigma[j, ] %*% u - xi0[j]
    if (u[j] < 0 && abs(value - lambda) > tol) {
      cat("Condition not met for u[", j, "] < 0\n")
      condition_met <- FALSE
    }
    if (u[j] == 0 && (value < -lambda - tol || value > lambda + tol)) {
      cat("Condition not met for u[", j, "] == 0\n")
      condition_met <- FALSE
    }
    if (u[j] > 0 && abs(value + lambda) > tol) {
      cat("Condition not met for u[", j, "] > 0\n")
      condition_met <- FALSE
    }
  }
  if (condition_met) {
    cat("All conditions are met.\n")
  } else {
    cat("Some conditions are not met.\n")
  }
}

# Check the stationary condition for the result
check_stationary_condition(result$u, Sigma_hat, xi0, lambda)
```
Since we have confirmed that the stationary condition is satisfied with a tolerance of $10^{-5}$, we can proceed with reporting the results for different $\tau$ values. Below is the translated code with comments in English:

5.
```{r}
#----------------------------------
library(MASS)
set.seed(1)

### training data
n <- 600
p <- 600

# generate Sigma
rho <- 0.3
Sigma <- matrix(rep(0.3, p^2), nrow = p, ncol = p)
diag(Sigma) <- 1
# generate matrix X and vector xi0 (xi_star in the above description)
X <- mvrnorm(n, rep(0,p), Sigma = Sigma)
xi0 <- Sigma %*% c(-1,1,-1,1,1,rep(0, p-5))
# Calculate Sigma hat (X^T X / n)
Sigma_hat <- t(X) %*% X / n

# Define the soft-thresholding function
soft_threshold <- function(z, gamma) {
  sign(z) * pmax(abs(z) - gamma, 0)
}

# Define the Proximal Gradient Descent function
proximal_gradient_descent <- function(Sigma, xi0, lambda, tau, delta = 1e-5, max_iter = 11000) {
  p <- ncol(Sigma)
  u <- rep(0, p)  # Initial value is 0
  iter_count <- 0
  u_history <- matrix(0, nrow = max_iter, ncol = p)
  
  for (iter in 1:max_iter) {
    u_old <- u
    gradient <- Sigma %*% u - xi0
    u <- soft_threshold(u - tau * gradient, tau * lambda)
    
    # Save the current iteration result
    u_history[iter, ] <- u
    
    # Check convergence criteria
    if (sum(abs(u - u_old), na.rm = TRUE) < delta) {
      iter_count <- iter
      u_history <- u_history[1:iter, ]  # Truncate to the actual number of iterations
      break
    }
  }
  
  if (iter == max_iter) {
    iter_count <- max_iter
  }
  
  list(u = u, iter_count = iter_count, u_history = u_history)
}

# Set parameters
lambda <- 1.2 * sqrt(log(p) / n)
tau_values <- c(0.01, 0.005, 0.001)

# Run experiments and report results
results <- list()
for (tau in tau_values) {
  cat("\nRunning Proximal Gradient Descent with tau =", tau, "\n")
  result <- proximal_gradient_descent(Sigma_hat, xi0, lambda, tau)
  results[[as.character(tau)]] <- result
  
  # Print results
  cat("Number of iterations:", result$iter_count, "\n")
  cat("Final iterate (first 10 components):", result$u[1:10], "\n")
}

# Function to plot the iterations of specific components
plot_iterations <- function(u_history, components, tau) {
  iter_count <- nrow(u_history)
  par(mfrow = c(length(components), 1), mar = c(4, 4, 2, 1))
  
  for (j in components) {
    plot(1:iter_count, u_history[, j], type = "l", main = paste("Component", j, "with tau =", tau),
         xlab = "Iteration", ylab = paste("u[", j, "]", sep = ""))
  }
}

# Plot iterations for components j = 2, 4, 6, 8, 10
for (tau in tau_values) {
  plot_iterations(results[[as.character(tau)]]$u_history, c(2, 4, 6, 8, 10), tau)
}

# Check the stationary condition with tolerance
check_stationary_condition <- function(u, Sigma, xi0, lambda, tol = 1e-3) {
  condition_met <- TRUE
  for (j in 1:length(u)) {
    value <- sum(Sigma[j, ] * u) - xi0[j]
    if (u[j] < 0 && abs(value - lambda) > tol) {
      cat("Condition not met for u[", j, "] < 0\n")
      condition_met <- FALSE
    }
    if (u[j] == 0 && (value < -lambda - tol || value > lambda + tol)) {
      cat("Condition not met for u[", j, "] == 0\n")
      condition_met <- FALSE
    }
    if (u[j] > 0 && abs(value + lambda) > tol) {
      cat("Condition not met for u[", j, "] > 0\n")
      condition_met <- FALSE
    }
  }
  if (condition_met) {
    cat("All conditions are met.\n")
  } else {
    cat("Some conditions are not met.\n")
  }
}

# Check the stationary condition for each result
for (tau in tau_values) {
  cat("\nChecking stationary condition for tau =", tau, "\n")
  check_stationary_condition(results[[as.character(tau)]]$u, Sigma_hat, xi0, lambda)
}
```
## Proximal Gradient Descent with Different $\tau$ Values

We performed the Proximal Gradient Descent algorithm using different values of $\tau$ and checked the stationary conditions. The tolerance for the stationary condition was set to $10^{-3}$. Here are the results for different $\tau$ values.

### Results

- For $\tau = 0.01$, the stationary condition is satisfied with a tolerance of $10^{-3}$.
- For $\tau = 0.005$, the stationary condition is satisfied with a tolerance of $10^{-3}$.
- For $\tau = 0.001$, the stationary condition requires a lower tolerance.

### Conclusion

For $\tau$ values of 0.01 and 0.005, the stationary condition is met with a tolerance of $10^{-3}$. However, for $\tau = 0.001$, the stationary condition requires a lower tolerance to be satisfied. This suggests that as the $\tau$ value decreases, a higher precision (lower tolerance) is necessary to meet the stationary condition.